Dominance Regions for Rank Two Cluster Algebras

### 2022-04-07 dominance_regions.tex
Small notation changes, some section 4 edits

  Important examples of pointed elements are the \emph{cluster monomials} of $\cA(b,c)$, i.e. the elements that can be written in the form $x_k^{a_1}x_{k+1}^{a_2}$ for some $k\in\ZZ$ and $a_1,a_2$ non-negative integers.
  Indeed, by [someresult of Qin] cluster monomials are part of any pointed basis in any cluster algebra.
  \say{Check if this needs full rank}

@@

  When $bc\le3$, the cluster algebra $\cA(b,c)$ will be of finite-type and cluster monomials form its only pointed basis.
  We will therefore assume that $bc\ge4$.
  In this case we give an explicit description of the dominance relation among $\bfg$-vectors.
  Specifically, we show that the $\bfg$-vector $\lambda$ dominates the collection of $\bfg$-vectors of the form $\lambda+(b a_1 ,c a_2)$, $a_1,a_2\in\ZZ$, inside its \emph{dominance polygon} $\cP(\lambda)$.
  %In this case, we compute the possible transitions between any two pointed bases.
  %The collection of $\bfg$-vectors which can appear in the deformation of a pointed basis is the \emph{dominance region}.
  \begin{theorem}
    \label{th:dominance inequalities}
    The dominance polygon $\cP(\lambda)$ pointed at $\lambda=(\lambda_1,\lambda_2)$ is the region defined by $x\leq \lambda_1, y\geq\lambda_2$, and the following inequalities:
    {
      \everymath={\displaystyle}
      \def\arraystretch{2.8}
      \[
        \begin{array}{rcccl}
          0 & \leq & \frac{b c-\sqrt{b c (b c-4)}}{2 b}(x-\lambda_1)+(y-\lambda_2) & \leq & \max\left(-c\lambda_1-\frac{b c+\sqrt{b c (b c-4)}}{2b}b\lambda_2,0\right)
          \\
          \min\left(-c\lambda_1,0\right) & \leq & \frac{b c-\sqrt{b c (b c-4)}}{2 b}(x-\lambda_1)-(y-\lambda_2) & \leq & 0
          \\
          0 & \leq &  -(x-\lambda_1)-\frac{b c-\sqrt{b c (b c-4)}}{2 c}(y-\lambda_2) & \leq & \max\left(\frac{b c+\sqrt{b c (b c-4)}}{2c}c\lambda_1+b\lambda_2,0\right)
          \\
          \min\left(b \lambda_2,0\right) & \leq & (x-\lambda_1) - \frac{b c-\sqrt{b c (b c-4)}}{2 c} (y-\lambda_2) & \leq & 0
        \end{array}
      \]
    }
  \end{theorem}

@@

  Here we explicitly compute the dominance regions $\cP(\lambda)$.
  When $\lambda$ is not imaginary, the dominance region is particularly simple.
  \begin{lemma}
    If $\lambda\in\ZZ^2\setminus\cI$, then $\cP(\lambda)=\{\lambda\}$.
  \end{lemma}
  \begin{proof}
    Any such $\lambda$ is the $\bfg$-vector of a cluster monomial, say $x_k^{a_1}x_{k+1}^{a_2}$.
    In this case the intersection
    \[\phi_{k-1}^{-1}\cC_{k-1}(\phi_{k-1}\lambda) \cap \phi_{k+1}^{-1}\cC_{k+1}(\phi_{k+1}\lambda)\]
    is precisely $\{\lambda\}$. 
    Indeed, when $k$ is odd the cone $\cC_{k-1}(\phi_{k-1}\lambda)$ lies entirely within a domain of linearity for $\phi_{k-1}^{-1}$.
    In particular, $\phi_{k-1}^{-1}\cC_{k-1}(\phi_{k-1}\lambda)$ is a cone containing $\lambda$ directed away from the origin (with walls parallel to the boundary of the domain of linearity containing $\lambda$).
    On the contrary, when $k$ is odd the cone $\cC_{k+1}(\phi_{k+1}\lambda)$ intersects the domain of linearity for $\phi_{k+1}^{-1}$ containing $\phi_{k+1}\lambda$ (resp. $\phi_{k-1}\lambda$) in a (possibly degenerate) convex quadrilateral with corners at $(0,0)$ and $\phi_{k+1}\lambda$ (resp. $\phi_{k-1}\lambda$).
    In particular, $\phi_{k+1}^{-1}\cC_{k+1}(\phi_{k+1}\lambda)$ (resp. $\phi_{k-1}^{-1}\cC_{k-1}(\phi_{k-1}\lambda)$) is a (possibly degenerate) convex quadrilateral with corners at $(0,0)$ and $\lambda$.
    The result follows.
  \end{proof}
  To compare with Theorem~\ref{th:dominance inequalities}, we consider three cases in which $\lambda\in\ZZ^2\setminus\cI$:
  \begin{itemize}
    \item if $\lambda_1 \le 0$ and $\lambda_2 \ge 0$, the second and fourth inequalities become equations for lines which intersect at $\lambda$;
    \item if $\lambda_1 > 0$ and $\lambda_2 \ge \frac{-bc+\sqrt{bc(bc-4)}}{2b}\lambda_1$, the first and second inequalities become equations for lines which intersect at $\lambda$;
    \item if $\lambda_2 < 0$ and $\lambda_1 \le \frac{bc-\sqrt{bc(bc-4)}}{2c}\lambda_2$, the third and fourth inequalities become equations for lines which intersect at $\lambda$.
  \end{itemize}
  This proves Theorem~\ref{th:dominance inequalities} when $\lambda\notin\cI$.

@@

  \begin{lemma}
    For $k\ge1$ and $\lambda\in\cI$, the region $\phi_k^{-1}\cC_k(\phi_k\lambda)\subset\RR^2$ is the polytope determined by the following inequalities:
    \begin{align}
      \label{ineq:1} u_{i,-}x+u_{i-1,+}y &\le u_{i,-}\lambda_1+u_{i-1,+}\lambda_2;\\
      \label{ineq:2} u_{i+1,-}x+u_{i,+}y &\le u_{i+1,-}\lambda_1+u_{i,+}\lambda_2;\\
      \label{ineq:3} -u_{i-1,-}x+u_{i,+}y &\le u_{i+1,-}\lambda_1+u_{i,+}\lambda_2;\\
      \label{ineq:4} -u_{i-1,-}x-u_{i-2,+}y &\le u_{i+1,-}\lambda_1+u_{i,+}\lambda_2.
    \end{align}
    \begin{align}
      \label{ineq:1} u_{k,-}(x-\lambda_1)+u_{k-1,+}(y-\lambda_2) &\le 0;\\
      \label{ineq:2} u_{k+1,-}(x-\lambda_1)+u_{k,+}(y-\lambda_2) &\le 0;\\
      \label{ineq:3} -u_{k-1,-}(x-\lambda_1)+u_{k,+}(y-\lambda_2) &\le cu_{k,+}\lambda_1;\\
      \label{ineq:4} -u_{k-1,-}(x-\lambda_1)-u_{k-2,+}(y-\lambda_2) &\le cu_{k,+}\lambda_1+bu_{k-1,-}\lambda_2.
    \end{align}

@@

@@

    \subsubsection*{Claim:} For $i\ge 0$, $\phi_{2i+1}^{-1}\cC_{2j+1}(\phi_{2j+1}\lambda)$ is the region determined by the inequalities 
    \begin{align*}
      \tag{a$'$} u_{2i+1,-}x+u_{2i,+}y &\le u_{2j+1,-}\lambda_1+u_{2j,+}\lambda_2;\\
      \tag{b$'$} u_{2i+2,-}x+u_{2i+1,+}y &\le u_{2j+2,-}\lambda_1+u_{2j+1,+}\lambda_2;\\
      \tag{c$'$} -u_{2i,-}x+u_{2i+1,+}y &\le u_{2j+2,-}\lambda_1+u_{2j+1,+}\lambda_2;\\
      \tag{d$'$} -u_{2i,-}x-u_{2i-1,+}y &\le u_{2j+2,-}\lambda_1+u_{2j+1,+}\lambda_2.
    \end{align*}
    By essentially the same calculations as above, these inequalities reproduce under the action of $\phi_2^{-1}$ and this completes the proof.
  \end{proof}

>>>>

  Important examples of pointed elements are the \emph{cluster monomials} of $\cA(b,c)$, i.e. the elements that can be written in the form $x_k^{a_k}x_{k+1}^{a_{k+1}}$ for some $k\in\ZZ$ and $a_k,a_{k+1}$ non-negative integers.
  Indeed, by [someresult of Qin] cluster monomials are part of any pointed basis in any cluster algebra.
  \say{Check if this needs full rank}

@@

  When $bc\le3$, the cluster algebra $\cA(b,c)$ will be of finite-type and cluster monomials form its only pointed basis.
  We will therefore assume that $bc\ge4$.
  In this case we give an explicit description of the dominance relation among $\bfg$-vectors.
  Specifically, we show that the $\bfg$-vector $\lambda$ dominates the collection of $\bfg$-vectors of the form $\lambda+(b a_0 ,c a_1)$, $a_0,a_1\in\ZZ$, inside its \emph{dominance polygon} $\cP(\lambda)$.
  %In this case, we compute the possible transitions between any two pointed bases.
  %The collection of $\bfg$-vectors which can appear in the deformation of a pointed basis is the \emph{dominance region}.
  \begin{theorem}
    \label{th:dominance inequalities}
    The dominance polygon $\cP(\lambda)$ pointed at $\lambda=(\lambda_1,\lambda_2)$ is the region defined by $x\leq \lambda_1, y\geq\lambda_2$, and the following inequalities:
    {
      \everymath={\displaystyle}
      \def\arraystretch{2.8}
      \[
        \begin{array}{rcccl}
          0 & \leq & \frac{b c-\sqrt{b c (b c-4)}}{2 b}(x-\lambda_1)+(y-\lambda_2) & \leq & \max\left(-c\lambda_1-\frac{b c+\sqrt{b c (b c-4)}}{2b}b\lambda_2,0\right)
          \\
          \min\left(-c\lambda_1,0\right) & \leq & \frac{b c-\sqrt{b c (b c-4)}}{2 b}(x-\lambda_1)-(y-\lambda_2) & \leq & 0
          \\
          0 & \leq &  -(x-\lambda_1)-\frac{b c-\sqrt{b c (b c-4)}}{2 c}(y-\lambda_2) & \leq & \max\left(\frac{b c+\sqrt{b c (b c-4)}}{2c}c\lambda_1+b\lambda_2,0\right)
          \\
          \min\left(b \lambda_2,0\right) & \leq & (x-\lambda_1) - \frac{b c-\sqrt{b c (b c-4)}}{2 c} (y-\lambda_2) & \leq & 0
        \end{array}
      \]
    }
  \end{theorem}

@@

  Here we explicitly compute the dominance regions $\cP(\lambda)$.
  When $\lambda$ is not imaginary, the dominance region is particularly simple.
  \begin{lemma}
    If $\lambda\in\ZZ^2\setminus\cI$, then $\cP(\lambda)=\{\lambda\}$.
  \end{lemma}
  \begin{proof}
    Any such $\lambda$ is the $\bfg$-vector of a cluster monomial, say $x_k^{a_k}x_{k+1}^{a_{k+1}}$.
    In this case the intersection
    \[\phi_{k-1}^{-1}\cC_{k-1}(\phi_{k-1}\lambda) \cap \phi_{k+1}^{-1}\cC_{k+1}(\phi_{k+1}\lambda)\]
    is precisely $\{\lambda\}$. 
    Indeed, when $k$ is odd the cone $\cC_{k-1}(\phi_{k-1}\lambda)$ lies entirely within a domain of linearity for $\phi_{k-1}^{-1}$.
    In particular, $\phi_{k-1}^{-1}\cC_{k-1}(\phi_{k-1}\lambda)$ is a cone containing $\lambda$ directed away from the origin (with walls parallel to the boundary of the domain of linearity containing $\lambda$).
    On the contrary, when $k$ is odd the cone $\cC_{k+1}(\phi_{k+1}\lambda)$ intersects the domain of linearity for $\phi_{k+1}^{-1}$ containing $\phi_{k+1}\lambda$ (resp. $\phi_{k-1}\lambda$) in a (possibly degenerate) convex quadrilateral with corners at $(0,0)$ and $\phi_{k+1}\lambda$ (resp. $\phi_{k-1}\lambda$).
    In particular, $\phi_{k+1}^{-1}\cC_{k+1}(\phi_{k+1}\lambda)$ (resp. $\phi_{k-1}^{-1}\cC_{k-1}(\phi_{k-1}\lambda)$) is a (possibly degenerate) convex quadrilateral with corners at $(0,0)$ and $\lambda$.
    The result follows.
  \end{proof}
  To compare with Theorem~\ref{th:dominance inequalities}, we consider three cases in which $\lambda\in\ZZ^2\setminus\cI$:
  \begin{itemize}
    \item if $\lambda_1 \le 0$ and $\lambda_2 \ge 0$, the second and fourth inequalities become equations for lines which intersect at $\lambda$;
    \item if $\lambda_1 > 0$ and $\lambda_2 \ge \frac{-bc+\sqrt{bc(bc-4)}}{2b}\lambda_1$, the first and second inequalities become equations for lines which intersect at $\lambda$;
    \item if $\lambda_2 < 0$ and $\lambda_1 \le \frac{bc-\sqrt{bc(bc-4)}}{2c}\lambda_2$, the third and fourth inequalities become equations for lines which intersect at $\lambda$.
  \end{itemize}
  This proves Theorem~\ref{th:dominance inequalities} when $\lambda\notin\cI$.

@@

  \begin{lemma}
    For $k\ge1$ and $\lambda\in\cI$, the region $\phi_k^{-1}\cC_k(\phi_k\lambda)\subset\RR^2$ is the polytope determined by the following inequalities:
    \begin{align}
      \label{ineq:1} u_{k,-}x+u_{k-1,+}y &\le u_{k,-}\lambda_1+u_{k-1,+}\lambda_2;\\
      \label{ineq:2} u_{k+1,-}x+u_{k,+}y &\le u_{k+1,-}\lambda_1+u_{k,+}\lambda_2;\\
      \label{ineq:3} -u_{k-1,-}x+u_{k,+}y &\le u_{k+1,-}\lambda_1+u_{k,+}\lambda_2;\\
      \label{ineq:4} -u_{k-1,-}x-u_{k-2,+}y &\le u_{k+1,-}\lambda_1+u_{k,+}\lambda_2.
    \end{align}

@@

    For $k=2j$, $j>0$, and $\lambda\in\cI$, we observe that $\cC_k(\phi_k\lambda)\subset\RR^2$ is given by the inequalities 
    \[ x - u_{2j+1,+}\lambda_1+u_{2j,+}\lambda_2 \le 0 \qquad\text{and}\qquad 0 \le y + u_{2j,-}\lambda_1+u_{2j-1,-}\lambda_2. \]
    \[ (\dagger)\ x\le u_{2j+1,+}\lambda_1+u_{2j,+}\lambda_2 \qquad\text{and}\qquad (\ddagger)\ -y\le u_{2j,-}\lambda_1+u_{2j-1,-}\lambda_2. \]
    We inductively compute the region $\phi_{2j}^{-1}\cC_{2j}(\phi_{2j}\lambda)$, using Lemma~\ref{le:transformed inequalities} and the equality $\phi_{2j}^{-1}=\phi_2^{-j}$.

@@

    \subsubsection*{Claim:} For $i\ge 0$, $\phi_{2i+1}^{-1}\cC_{2j+1}(\phi_{2j+1}\lambda)$ is the region determined by the inequalities 
    \begin{align*}
      \tag{a$'$} u_{2i+1,-}x+u_{2i,+}y &\le u_{2j+1,-}\lambda_1+u_{2j,+}\lambda_2;\\
      \tag{b$'$} u_{2i+2,-}x+u_{2i+1,+}y &\le u_{2j+2,-}\lambda_1+u_{2j+1,+}\lambda_2;\\
      \tag{c$'$} -u_{2i,-}x+u_{2i+1,+}y &\le u_{2j+2,-}\lambda_1+u_{2j+1,+}\lambda_2;\\
      \tag{d$'$} -u_{2i,-}x-u_{2i-1,+}y &\le u_{2j+2,-}\lambda_1+u_{2j+1,+}\lambda_2.
    \end{align*}
    By essentially the same calculations as above, these inequalities reproduce under the action of $\phi_2^{-1}$ and this completes the proof.
  \end{proof}
  The following consequence is easily seen from the Chebyshev recursion.
  \begin{corollary}
    \begin{align*}
      u_{k,-}(x-\lambda_1)+u_{k-1,+}(y-\lambda_2) &\le 0;\\
      u_{k+1,-}(x-\lambda_1)+u_{k,+}(y-\lambda_2) &\le 0;\\
      -u_{k-1,-}(x-\lambda_1)+u_{k,+}(y-\lambda_2) &\le cu_{k,+}\lambda_1;\\
      -u_{k-1,-}(x-\lambda_1)-u_{k-2,+}(y-\lambda_2) &\le cu_{k,+}\lambda_1+bu_{k-1,-}\lambda_2.
    \end{align*}
  \end{corollary}
